---
title: "Graph clustering with the Stochastic block model"
author: "Etienne Côme"
date: "`r Sys.Date()`"
fig_caption: yes
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Graph clustering with the Stochastic block model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup-knitr, include = FALSE, message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


Loads packages and set a future plan for parallel processing.
```{r setup-packages,message=FALSE, include=TRUE}
library(future)
library(Matrix)
library(ggplot2)
library(greed)
library(dplyr)
library(ggpubr)
```


```{r ggplot-theme, include=FALSE, echo=FALSE, message=FALSE}
source("./vignettes_ggtheme.R")
```

# The model

# An introductory example: SBM with hierarchicaltructure

We begin by simulating from a hierarchically structured SBM model, with 3 large clusters, each composed of 5 smaller clusters with higher connection probabilities.
```{r, fig.show='hold'}
N=400
K=6
pi=rep(1/K,K)
lambda  = 0.1
lambda_o = 0.01
Ks=3
mu = bdiag(lapply(1:(K/Ks), function(k){
  matrix(lambda_o,Ks,Ks)+diag(rep(lambda,Ks))}))+0.001
sbm = rsbm(N,pi,mu)
```

```{r}
df = expand.grid(x=1:K,y=1:K)
df$Theta=c(t(matrix(mu)))
leg_title = TeX(r'($\theta_{kl}$)')
leg_breaks = c(1e-3, 0.03, 0.13)
gg = ggplot(df, aes(x = x, y = y, fill=Theta, alpha=Theta)) +
  geom_tile() +
  xlab('l') +
  ylab('k') +
  ggplot2::scale_fill_distiller(leg_title, breaks = leg_breaks, palette="YlOrRd",direction = 1,guide = ggplot2::guide_legend(),limits=c(0,max(leg_breaks))) +
  ggplot2::scale_alpha(leg_title,range=c(0,1),limits=c(0,max(leg_breaks)), breaks = leg_breaks) +
  ggplot2::coord_fixed()+ggplot2::theme_bw() +
  theme(
    panel.background = element_rect(fill = "transparent"), # bg of the panel
    plot.background = element_rect(fill = "transparent", color = NA), # bg of the plot
    panel.grid.major = element_blank(), # get rid of major grid
    panel.grid.minor = element_blank(), # get rid of minor grid
    legend.background = element_rect(fill = "transparent"), # get rid of legend bg
    legend.box.background = element_rect(fill = "transparent"), # get rid of legend panel bg
    text = element_text(size=20)
  )
```


Perform the clustering with default  algorithm. We specify to choose an sbm model since for squared sparse matrix the default is a DcSbm model. An hybrid  algorithm is selected by default and the default value for the parameter K is 20.

```{r}
sol = greed(sbm$x,model = Sbm())
```

Look at the parameters estimate:

```{r}
coef(sol)
```

Plot the results using a block representation.

```{r, fig.show='hold',out.width="70%",fig.width=8,fig.height=5.5}
plot(sol,type='blocks')
```



Plot the results with a node link diagram.
```{r, fig.show='hold',out.width="70%",fig.width=8,fig.height=5.5}
plot(sol,type='nodelink')
```

Or a dendrogram for selecting a smaller value for K.
```{r, fig.show='hold',out.width="70%",fig.width=8,fig.height=5.5}
plot(sol,type='tree')
```

Eventually study the evolution of $-log(\alpha)$ with respect to $K$.
```{r, fig.show='hold',out.width="70%",fig.width=8,fig.height=5.5}
plot(sol,type='path')
```

Or of ICL with respect to $log(\alpha)$
```{r, fig.show='hold',out.width="70%",fig.width=8,fig.height=5.5}
plot(sol,type='front')
```

And select a smaller value to extract a new solution.
```{r, fig.show='hold',out.width="70%",fig.width=8,fig.height=5.5}
sol2 = cut(sol,2)
plot(sol2,type='blocks')
```

We may try with a dcsbm simulation and compare models with and without degree correction:

```{r}
sim_dcsbm <- rdcsbm(N,pi,mu,round(rexp(N,1/15)),round(rexp(N,1/15)))
X <- sim_dcsbm$x
X[X>1] <- 1
sol_dcsbm <- greed(X,model = DcSbm())
sol_dcsbm
sol_sbm <- greed(X,model = Sbm())
sol_sbm
```

As expected the degree corrected version did a better job as the ICL's suggest. Without degree correction the model has to use more groups to fit the degree heterogeneity. 

## Some simple real data, the books dataset

```{r}
data(Books)
sol_dcsbm = greed(Books$X,model = DcSbm())
sol_dcsbm
sol_sbm = greed(Books$X,model = Sbm())
sol_sbm
```

The network as been well recognized as an undirected graph. We may look at the fitted model prior to check:

```{r}
prior(sol_dcsbm)
```

For this dataset the model without degree correction seem's to reach a better solution than its concurrent.

```{r, fig.show='hold',out.width="100%",fig.width=8,fig.height=5.5}
bl_sbm = plot(sol_sbm,type='blocks')
bl_dcsbm = plot(sol_dcsbm,type='blocks')
ggarrange(bl_sbm,bl_dcsbm)
```

```{r, fig.show='hold',out.width="70%",fig.width=8,fig.height=5.5}
plot(sol_sbm,type='tree')
```

```{r, fig.show='hold',out.width="70%",fig.width=8,fig.height=5.5}
library(ggraph)
library(tidygraph)

graph <- igraph::graph_from_adjacency_matrix(Books$X) %>% as_tbl_graph() %>% 
    mutate(Popularity = centrality_degree())  %>% 
    activate(nodes) %>%
    mutate(cluster=factor(clustering(sol_sbm),1:K(sol_sbm)))

# plot using ggraph
ggraph(graph, layout = 'kk') + 
    geom_edge_link() + 
    geom_node_point(aes(size = Popularity,color=cluster))

```

```{r}
sol_sbm_k3 = cut(sol_sbm,3)
table(clustering(sol_sbm_k3),Books$label)
```

## Multinomial SBM, layered networks 

#### Newguinea dataset
```{r}
data("NewGuinea")
dim(NewGuinea)
```

```{r}
sol_newguinea = greed(NewGuinea,model=MultSbm())
```

```{r, fig.show='hold',out.width="100%",fig.width=8,fig.height=5.5}
plot(sol_newguinea,type='blocks')
```


```{r, fig.show='hold',out.width="70%",fig.width=8,fig.height=5.5}

enemies = as.data.frame(which(NewGuinea[,,1]==1,arr.ind = TRUE)) %>% 
  mutate(relation="enemy")
friends = as.data.frame(which(NewGuinea[,,2]==1,arr.ind = TRUE)) %>%
  mutate(relation="friend")

edges = rbind(enemies,friends)
graph = tbl_graph(edges=edges,nodes=data.frame(id=1:16,cluster=factor(clustering(sol_newguinea))))                          
                          
# plot using ggraph
ggraph(graph, layout = 'kk', weights = if_else(relation=="friends",1,0)) + 
    geom_edge_link(aes(color=relation),width=1.2) + 
    scale_edge_color_manual(values = c("friend"="#229922","enemy"="#992222"))+
    geom_node_point(aes(shape=cluster),size=5)

```



```{r}
mod <- MixedModels(list(pos = DcSbmPrior(), neg = DcSbmPrior()))
data <- list(pos = NewGuinea[, , 2], neg = NewGuinea[, , 1])
sol_bidcsbm <- greed(data, model = mod, K = 5)
```

```{r, fig.show='hold',out.width="100%",fig.width=8,fig.height=5.5}
bl_pos = plot(extractSubModel(sol_bidcsbm,"pos"),type="blocks")
bl_neg = plot(extractSubModel(sol_bidcsbm,"neg"),type="blocks")
ggarrange(bl_pos,bl_neg)
```


#### 7th grade dataset


```{r}
data("SevenGraders")
dim(SevenGraders)
```

```{r}
mod <- MixedModels(list(class = DcSbmPrior(), friends = DcSbmPrior(), work = DcSbmPrior()))
data <- list(class = SevenGraders[, , 1], friends = SevenGraders[, , 2], work = SevenGraders[, , 3])
sol <- greed(data, model = mod, K = 5)
```

```{r, fig.show='hold',out.width="70%",fig.width=8,fig.height=5.5}
plot(extractSubModel(sol, "work"), type = "blocks")
```

## bi-partite graph



